Association between trauma severity models and opportunities for improvement:

### 2022-12-20 manuscript.Rmd
revised paper according to feedbck, wrote abstract

---
title: 'Association between trauma severity models and opportunities for improvement:
  A retrospective cohort study'
author: "Victoria Witthoff"
output:
  pdf_document: default
  html_document:
    df_print: paged
    number_sections: true
  word_document: default
bibliography: bibliography.bib
csl: vancouver.csl
fontsize: 11pts
linestretch: 1
always_allow_html: yes
editor_options: 
   chunk_output_type: inline
---

@@

```{r main}
  source("main.R") 
```

@@

<!-- Provide in the abstract an informative and balanced summary of
what was done and what was found. Not more than 300 words. -->

@@

Introduction
============
Trauma, defined as the clinical entity composed of physical injury and the body's associated response [@Gerdin2015], is one of the leading causes of death worldwide and accounts for approximately 9% of global deaths annually [@globalburden]. The global health burden of trauma has led to it being described as a worldwide pandemic. One of the characteristics of this particular pandemic is its demographic spread; trauma is a leading cause of death for people aged 10-49 years [@globalburden] and 90% of trauma-related deaths occur in low- to middle-income countries [@whoepi]. Non-fatal outcomes also comprise a significant portion of the global health burden of disability [@globalburden], which can drastically reduce quality of life and health status in survivors of severe trauma [@globalburden; @longtermtrauma]. The quality of trauma care is therefore a globally relevant issue with significance for a great number of people. Improvements in trauma outcomes benefit both the individual and also society at large by reducing the societal and healthcare costs of disability. The significance of the consequences and burden of trauma has been recognised by the UN, who in their Sustainable Development Goals included the aim of reducing the number of global deaths and injuries from road traffic accidents by half by 2020 [@sdg]. The importance of improving trauma care named as one of the strategies in achieving this goal. The goal was not met by 2020 [@sdg20], which is another signal that there is significant progress to be made in trauma care.

Efforts to advance the state of trauma care have been ongoing for many years. Today many trauma centres have implemented quality improvement (QI) programs, and it has been shown in the US that these programs improve outcomes [@ustqi]. QI programs use many tools, including statistical analysis and audit filters [@hakkenbrakprevdeath], but the gold standard of trauma QI is the multidisciplinary morbidity and mortality (M&M) conference. These conferences review selected patient cases and establish whether there was opportunity for improvement (OFI) in each case. One method of improving the quality of M&M conferences is by improving case selection. This is usually achieved by using audit filters that automatically flag cases for review. One approach is the use of trauma severity models as audit filters. Trauma severity models are algorithms that evaluate trauma severity in quantitative terms. The effectiveness of these models in case selection is understudied and their potential as audit filters is the focus of this paper.

@@

The significance of M&M conferences in trauma QI is a consequence of the unique challenges inherent to trauma care. Management of severe trauma is highly time-sensitive and is also dependent on the actions of a chain of healthcare providers spanning multiple professions and disciplines. Outcomes in severe trauma are dependent on the quality of care received [@ustqi] and errors in management are a common cause of preventable deaths [@prevdeathsghorbani; @prevdeathsmontmany; @prevdeathsteixeira]. Since mistakes by any one provider can lead to severe outcomes, attempts to improve trauma outcomes must seek to improve all aspects of the entire chain of care. This can only be achieved through peer review and consequently, M&M conferences are the cornerstone of trauma QI programs.

As mentioned, M&M conferences review a selection of patient cases, attempt to establish whether there were opportunities for improvement (OFI) in each specific case, and, where necessary, implement corrective actions. The occurrence of OFI function in this manner as a proxy for incidence of suboptimal care. Regular conferences allow QI programs to identify patterns of errors and provide a basis for systemic quality improvement. To maximise effectiveness, conferences should be attended by all involved specialities and have a structured format. The WHO guidelines for trauma QI programs give specific recommendations as to the structure and format of M&M conferences [@whoguidelines]. Despite theserecommendations, there is a still a great degree of variation amongst clinics in how M&M conferences are carried out, with a significant number of clinics still using an informal and unstructured format [@tqipanglo]. The World Health Organisation (WHO) themselves note in their guidelines that M&M conferences are not used to their fullest potential globally [@whoguidelines].

@@

The use of trauma severity models as audit filters is not limited to solely mortalities; some hospitals also use them as audit filters in survivors. However, the extent to which trauma severity models are associated with or can predict OFI as found in M&M conferences is not well understood. The two studies mentioned investigated only preventable deaths and to date there is no published research that investigates whether trauma severity models can predict OFI across the board.

@@

This is a single-centre retrospective cohort study, using data from two registries at Karolinska University Hospital: the trauma registry and the trauma care quality database. The trauma care quality database is a subset of the trauma registry and contains data on trauma patients who were selected for M&M reviews. These two databases were linked and then analysed to estimate the association between the common trauma severity models TRISS, NORMIT and PS and OFI using logistic regression. The predictive performance of these models was evaluated using measures of discrimination and calibration.

@@

@@

The trauma registry at Karolinska University Hospital is part of SweTrau, Sweden's national trauma registry, and follows SweTrau guidelines. All trauma patients meeting the SweTrau inclusion criteria are registered in the trauma registry by research nurses at the trauma centre. Data is collected from the patient's electronic medical records. The trauma registry contains data on patients from 2013 to 2021. <!-- Based on the dates I see in the Swetrau dataset -->

@@

The trauma care quality database is an internal hospital registry and registers the outcome of M&M conferences. M&M conferences are held at Karolinska University Hospital ten times per year and are attended by all involved specialties and professions. Each case reviewed is registered in the trauma care quality register with data about the outcome of the review. The database contains data on patients from 2014 to 2021. <!-- Double check these dates -->

@@

The trauma care quality database contains patients from the trauma registry that were selected for M&M review. At Karolinska University Hospital, all fatalities are automatically selected for M&M review. In addition to this, the electronic medical records of all trauma patients are screened by research nursing staff to identify cases that potentially received sub-optimal care. These nurses also use audit filters that automatically highlight cases with abnormal parameters. Cases that two research nurses have evaluated and found to have potential for sub-optimal care are also selected for M&M review. Figure 1 gives an overview of the workflow that leads to determination of OFI.

```{r, echo=FALSE, out.width = "400px"}
knitr::include_graphics("images/ofi.png") ## ofi_diagram.R; save_png(ofi_diagram(), "ofi") to update
## titles for plots will be added in word
```

@@

### Outcome
The studied outcome is the binary variable "opportunities for improvement" (OFI), as identified by the multidisciplinary M&M conference at Karolinska University Hospital. OFI is registered in the trauma care quality database and is coded as either "Yes - at least one opportunity for improvement identified" or "No - no opportunities for improvement identified".

### Exposures
The exposures of interest are the trauma severity models TRISS, NORMIT, and PS. All three of these models estimate the probability of death, expressed as a probability. The scores given by each of these models were calculated using physiological and anatomical parameters recorded in the trauma registry. Conversion and handling of these variables was carried out according to the SweTrau manual.

@@

The algorithm for PS includes the patient's age, gender, GCS and ISS.

@@

R, a programming language for statistical computing, was used for data collection and analysis. Logistic regression was used to estimate association between the estimates of the trauma severity models and OFI and the effect size was given as the OR. The CI for the OR was computed using the Wald estimator. Receiver operator characteristic (ROC) curves were used to assess diagnostic performance. The area under the curve (AUC) was calculated to compare the models with one another. The Integrated Calibration Index (ICI) was used to measure calibration. Accuracy was plotted and is reported as the accuracy at a 0.5 cutoff. Confidence intervals were calculated with bootstrapping. Results were presented with a confidence interval of 95%. A p-value of <0.05 was considered significant.

@@

All patient information used in this study is pseudononymised and stored in a secure database. Patient medical records will not be accessed. Patients were not asked to give consent for inclusion in the Karolinska University Hospital trauma registry. The ability and ethical approval to collect patient data for this registry without pre-emptively asking for patient consent is well-established, as collecting this data is deemed to be in the public interest according to article 6.1 e of GDPR regulations. Collecting data is in the public interest due to the fact that this patient register is used to improve management of trauma, which is a leading cause of death. Patients have the right to be removed from the register on request. They also have the right to a copy of data that is stored about them on request. <!-- https://meetings.lg.se/pdfFile/pdfFile.aspx?objectID=%2F7Iqgtt89nW2zhRjMSNZ%2Fw%3D%3D -->

This project is a retrospective cohort study and the project has thereby had no effect on patient treatment. There is no risk that patients are harmed by inclusion in the study. This study has the potential to improve patient care by analysing a potential factor that could predict suboptimal trauma care. It is true that the patient population for this study includes vulnerable populations, such as the deceased and the severely injured, but this is justified by the fact that this population stands to gain the most from improved trauma care.

@@

Participants
------------
A total of `r exclusion[1,2]` patients were recorded in the trauma registry during the study period. Of these, `r exclusion[2,2]` were reviewed at an M&M conference and therefore became eligible for study inclusion. `r exclusion[7,3]` patients were excluded, due to either being of paediatric age (`r exclusion[3,3]`), being dead on arrival at hospital (`r exclusion[4,3]`), or due to incomplete data (`r exclusion[5,3]`). Figure 2 shows the inclusion process for eligible participants.

```{r, echo=FALSE, out.width = "400px"}
knitr::include_graphics("images/exclusion.png") ## exclusion_flowchart.R; save_png(exclusion_flowchart(exclusion), "exclusion") to update
```

Detailed totals of missing data for each required parameter in cases that were excluded for incomplete data are given in table 1. Some patients had missing data in several parameters and therefore the total number of missing parameters is greater than the total number of patients excluded due to incomplete data.

```{r}
 as_latex(table(na.data, type = "incomplete"))
```

Descriptive data
----------------
A total of `r exclusion[6,2]` participants were identified as eligible for inclusion. Their demographical and clinical characteristics are described in table 2.

@@

```{r} 
  table.one
```

Main results
------------

The fitted curves for the logistic regression models of each score are shown in figure 3. The odds ratios for TRISS, NORMIT and PS were `r t[['or']]` (95% CI: `r t[['or.ci.lo']]`-`r t[['or.ci.hi']]`, *p*: `r t[['or.p']]`), `r n[['or']]` (95% CI: `r n[['or.ci.lo']]`-`r n[['or.ci.hi']]`, *p*: `r n[['or.p']]`), and `r p[['or']]` (95% CI: `r p[['or.ci.lo']]`-`r p[['or.ci.hi']]`, *p*: `r p[['or.p']]`) respectively.

```{r}
knitr::include_graphics("images/glm.png") ## update: glm_x3(df)
```

The receiver operating curves for all scores are plotted in 4. The areas under the curve for TRISS, NORMIT and PS were `r t[['auc']]` (95% CI: `r t[['auc.ci.lo']]`-`r t[['auc.ci.hi']]`), `r n[['auc']]` (95% CI: `r n[['auc.ci.lo']]`-`r n[['auc.ci.hi']]`), and `r p[['auc']]` (95% CI: `r p[['auc.ci.lo']]`-`r p[['auc.ci.hi']]`) respectively.

```{r}
knitr::include_graphics("images/roc.png") ## update: roc_x3(t, n, p)
```

The integrated calibration index for TRISS, NORMIT and PS was `r t[['ici']]` (95% CI: `r t[['ici.ci']][1]`-`r t[['ici.ci']][2]`, `r n[['ici']]` (95% CI: `r n[['ici.ci']][1]`-`r n[['ici.ci']][2]` and `r p[['ici']]` (95% CI: `r p[['ici.ci']][1]`-`r p[['ici.ci']][2]` respectively. The accuracy at all cutoffs is plotted for all models in  5. The accuracy at a cutoff of 0.5 for TRISS, NORMIT and PS was `r t[['acc']]` (95% CI: `r t[['acc.ci']][1]`-`r t[['acc.ci']][2]`), `r n[['acc']]` (95% CI: `r n[['acc.ci']][1]`-`r n[['acc.ci']][2]`) and `r p[['acc']]` (95% CI: `r p[['acc.ci']][1]`-`r p[['acc.ci']][2]`) respectively.

```{r}
knitr::include_graphics("images/acc.png") ## update: acc_x3(t, n, p)
```

The differences between the scores for each result are shown in table x.

```{r}
 as_latex(table(d, type = "deltas"))
```

@@

The results of this study agree with previous studies assessing the predictive power of trauma severity models with regards to preventable deaths. Although this study was not specifically about preventable deaths, the factor that determines whether a death was preventable is whether an improvement in care could have prevented the death, i.e. whether there was a significant opportunity for improvement, which makes the previous studies on preventable deaths broadly relevant to this study. Ghorbani et al found that both TRISS and NORMIT had poor performance in predicting preventable deaths [@prevdeathsghorbani]. Heim et al found that PS and TRISS miss a significant proportion of preventable deaths [@brohi16]. Our results also show that none of the three models investigated (TRISS, NORMIT and PS) were able to reliably predict OFI across the board and not only in mortalities. The difference between our study and prior studies is that we looked at the trauma population across the board and not only deaths.

Even though our results were expected, it was still important to investigate this topic scientifically because trauma severity models are in use today as audit filters. Our findings show that the use of trauma severity models as audit filters is unlikely to lead to improvements in case selection, and in fact could be harmful. Audit filters that don't improve case selection will either lead to cases being over-selected, causing resource wastage, or to cases being under-selected, leading to OFI being missed. Trauma severity models are useful when used on the population level, to allow institutions to compare themselves and to provide a framework for discussing trauma severity. They are much less useful when applied to individuals, and were not intended for that use.

@@

### Health equity
The demographical characteristics of the participants in this study reflect the well-known demographics of trauma patients worlwide. `r dd['male']`% of participants were male and the median age was `r dd['age']`, which matches with statistics both in Sweden and globally [@globalburden]. The predominance of male patients invites one to question if female patients are less likely to receive quality care. In healthcare as in all areas, repetition leads to mastery of a skill, so an onlooker might be tempted to assume that the high volume of male patients in trauma will lead healthcare providers to superiority iin treating males, whilst treatment of female patients will be neglected. This demographical difference has precipitated research that seeks to investigate whether this presumption holds true, and several large multi-centre studies have been published that evaluated whether there is a correlation between sex and outcomes. None of these studies found a correlation beween sex and increased mortality or other serious outcomes [@croce2002; @magnotti2008; @pape2019; @trentzsch2014], and in fact one study in the Netherlands found that female sex was correlated with decreased mortality [@pape2019].

>>>>

---
title: 'Association between trauma severity models and opportunities for improvement:
  A retrospective cohort study'
author: "Victoria Witthoff"
output:
  bookdown::pdf_document2: default
  bookdown::html_document2: default
  bookdown::word_document2: default
bibliography: bibliography.bib
csl: vancouver.csl
fontsize: 11pts
linestretch: 1
always_allow_html: yes
editor_options: 
   chunk_output_type: inline
---

@@

```{r main}
  source("main.R") 
```
<!-- rmarkdown::render("manuscript.Rmd", html_document2(global_numbering = TRUE)) -->

@@

<!-- Provide in the abstract an informative and balanced summary of
what was done and what was found. Not more than 300 words. -->
*Background:* Morbidity and mortality review, which aims to discover opportunities for improvement (OFI) is a cornerstone of trauma quality improvement worldwide. The process of case selection and review is time-consuming, leading many hospitals to use audit filters to automate case selection. Trauma severity models, which estimate probability of survival, have been proposed as one such audit filter. The ability of these models to predict opportunities for improvement (OFI) is not well understood. Two studies have found that they perform poorly at predicting preventable deaths, but as of yet there is no published literature that analyses the performance of these models across the board and not only in deaths. *Aim:* To evaluate how the commonly used models TRISS, NORMIT and PS are associated with, and to what extent they can predict, OFI. *Methods:* This is a retrospective registry-based cohort study at Karolinska University Hospital covering the period 2014-2021. Logistic regression was used to estimate the association between the models and OFI. Receiver operating characteristic curves with the corresponding area under the curve (AUC) were used to assess discrimination and the Integrated Calibration Index was used to assess calibration. *Results:* 5320 participants were included from a seven-year period. Results were similar for all models.  Lower probability of survival was associated with higher probability of OFI, with TRISS having the strongest association (derived OR: `r t['or']`). Predictive performance and calibration was poor across the board. TRISS had the best AUC of `r t['auc']`. NORMIT had the best calibration with an ICI of `r n['ici']`. *Conclusions:* Although low probability of survival is associated with increased odds of OFI, none of the models were able to reliably predict OFI. The use of these models as audit filters in case selection for M&M conferences is not supported by our results and may lead to over-selection.

@@

Introduction
============
Trauma, defined as the clinical entity composed of physical injury and the body's associated response [@Gerdin2015], is one of the leading causes of death worldwide and accounts for approximately 9% of global deaths annually [@globalburden]. The global health burden of trauma has led to it being described as a worldwide pandemic. One of the characteristics of this particular pandemic is its demographic spread; trauma is a leading cause of death for people aged 10-49 years [@globalburden] and 90% of trauma-related deaths occur in low- to middle-income countries [@whoepi]. Non-fatal outcomes also comprise a significant portion of the global health burden of disability [@globalburden], which can drastically reduce quality of life and health status in survivors of severe trauma [@globalburden; @longtermtrauma].

The quality of trauma care is therefore a globally relevant issue with significance for a great number of people. Improvements in trauma outcomes benefit both the individual and also society at large by reducing the societal and healthcare costs of disability. The significance of the consequences and burden of trauma has been recognised by the UN, who in their Sustainable Development Goals included the aim of reducing the number of global deaths and injuries from road traffic accidents by half by 2020 [@sdg]. The importance of improving trauma care named as one of the strategies in achieving this goal. The goal was not met by 2020 [@sdg20], which is another signal that there is significant progress to be made in trauma care.

Efforts to advance the state of trauma care have been ongoing for many years. Today many trauma centres have implemented quality improvement (QI) programs, and it has been shown in the US that these programs improve outcomes [@ustqi]. QI programs use many tools, including statistical analysis and audit filters [@hakkenbrakprevdeath], but the gold standard of trauma QI is the multidisciplinary morbidity and mortality (M&M) conference. These conferences review selected patient cases and establish whether there was opportunity for improvement (OFI) in each case.

One area that is understudied is the use of trauma severity models as audit filters in case selection for M&M conferences. These models are algorithms that estimate the probability of survival. It has been proposed that using such models as audit filters to automatically flag cases for review could improve the effectiveness of M&M conferences while also reducing their resource burden [@brohi16; @prevdeathsghorbani ]. Many hospitals have already implemented this approach [@brohi16]; however, the ability of trauma severity models to predict OFI is not well understood and will be investigated in this paper.

@@

The significance of M&M conferences in trauma QI is a consequence of the unique challenges inherent to trauma care. Management of severe trauma is highly time-sensitive and is also dependent on the actions of a chain of healthcare providers spanning multiple professions and disciplines. Outcomes in severe trauma are dependent on the quality of care received and errors in management are a common cause of preventable deaths [@ustqi; @prevdeathsghorbani; @prevdeathsmontmany; @prevdeathsteixeira]. Since mistakes by any one provider can lead to severe outcomes, attempts to improve trauma outcomes must seek to improve all aspects of the entire chain of care. This can only be achieved through peer review and consequently, M&M conferences are the cornerstone of trauma QI programs.

The purpose of M&M conferences is to review selected patient cases, establish whether there were opportunities for improvement, and implement corrective action. The occurrence of OFI function in this manner as a proxy for incidence of suboptimal care. Regular conferences allow QI programs to identify patterns of errors and provide a basis for systemic quality improvement. To maximise effectiveness, conferences should be attended by all involved specialities and have a structured format. The WHO guidelines for trauma QI programs give specific recommendations as to the structure and format of M&M conferences [@whoguidelines]. Despite these recommendations, there is a still a great degree of variation amongst clinics in how M&M conferences are carried out, with a significant number of clinics still using an informal and unstructured format [@tqipanglo]. The World Health Organisation (WHO) themselves note in their guidelines that M&M conferences are not used to their fullest potential globally [@whoguidelines].

@@

The studies mentioned signify that trauma severity models are not well suited to predicting OFI in fatalities and indicate that they should not be used for this purpose. However, the extent to which trauma severity models are associated with or can predict OFI is not well understood, as research thus far has only analysed their role in predicting preventable deaths. To date there is no published research that investigates whether trauma severity models can predict OFI across the board.

@@

This is a single-centre retrospective cohort study, using data from two registries at Karolinska University Hospital: the trauma registry and the trauma care quality database. The trauma care quality database is a subset of the trauma registry and contains data on trauma patients who were selected for M&M reviews. These two databases were linked and then analysed to estimate the association between the common trauma severity models TRISS, NORMIT, and PS and OFI using logistic regression. The predictive performance of these models was evaluated using measures of discrimination and calibration.

@@

<!-- Describe the setting, locations, and relevant dates, including
periods of recruitment, exposure, follow-up, and data collection -->

@@

The trauma registry at Karolinska University Hospital is part of SweTrau, Sweden's national trauma registry, and thereby follows SweTrau guidelines. All trauma patients meeting the SweTrau inclusion criteria are registered in the trauma registry by research nurses at the trauma centre. Data is collected from the patient's electronic medical records.

@@

The trauma care quality database is an internal hospital registry and registers the outcome of M&M conferences. M&M conferences are held at Karolinska University Hospital ten times per year and are attended by all involved specialties and professions. Each case reviewed is registered in the trauma care quality register with data about the outcome of the review. The database contains data on patients from 2014 to 2021.

@@

The trauma care quality database contains patients from the trauma registry that were selected for M&M review. At Karolinska University Hospital, all fatalities are automatically selected for M&M review. In addition to this, the electronic medical records of all trauma patients are screened by research nursing staff to identify cases that potentially received sub-optimal care. These nurses also use audit filters that automatically highlight cases with abnormal parameters. Cases that two research nurses have evaluated and found to have potential for sub-optimal care are also selected for M&M review. Figure \@ref(fig:ofi) gives an overview of the workflow that leads to determination of OFI.

```{r ofi, fig.cap = "Workflow for case selection and determining OFI", fig.align = "center", echo=FALSE, out.width = "600px"}
knitr::include_graphics("images/ofi.png") 
## ofi_diagram.R; save_png(ofi_diagram(), "ofi") to update
```

@@

### Outcome
The studied outcome was the binary variable "opportunities for improvement" (OFI), as identified by the multidisciplinary M&M conference at Karolinska University Hospital. OFI is registered in the trauma care quality database and is coded as either "Yes - at least one opportunity for improvement identified" or "No - no opportunities for improvement identified".

### Exposures
The exposures of interest were the trauma severity models TRISS, NORMIT, and PS. All three of these models estimate the probability of death, expressed as a probability. The scores given by each of these models were calculated using physiological and anatomical parameters recorded in the trauma registry. Conversion and handling of these variables was carried out according to the SweTrau manual.

@@

The algorithm for PS includes the patient's age, gender, GCS and ISS. We chose to use the PS12 algorithm instead of the newer PS19 model due to our registry lacking data needed to calculate PS19.

@@

R, a programming language for statistical computing, was used for data collection and analysis [@r]. Logistic regression was used to estimate association between the estimates of the trauma severity models and OFI and the effect size was given as the OR. The CI for the OR was computed using the Wald estimator. Receiver operator characteristic (ROC) curves were used to assess diagnostic performance. The area under the curve (AUC) was calculated to compare the models with one another. The Integrated Calibration Index (ICI) was used to measure calibration. Accuracy was plotted and is reported as accuracy at a >=0.5 cutoff. Confidence intervals for the AUC, ICI and accuracy were computed with 1000 stratified bootstrap replicates. Results are presented with a confidence interval of 95%. A p-value of <0.05 was considered significant.

@@

All patient information used in this study is pseudononymised and stored in a secure database. Patient medical records will not be accessed. Patients were not asked to give consent for inclusion in the Karolinska University Hospital trauma registry. The ability and ethical approval to collect patient data for this registry without pre-emptively asking for patient consent is well-established, as collecting this data is deemed to be in the public interest according to article 6.1 e of GDPR regulations. Collecting data is in the public interest due to the fact that this patient register is used to improve management of trauma, which is a leading cause of death. Patients have the right to be removed from the register on request. They also have the right to a copy of data that is stored about them on request.

This project is a retrospective cohort study and the project has thereby had no effect on patient treatment. There is no risk that patients are harmed by inclusion in the study. This study has the potential to improve patient care by analysing a potential factor that could predict suboptimal trauma care. It is true that the patient population for this study includes vulnerable populations, such as the deceased and the severely injured, but this is justified in the opinion of this author by the fact that this population stands to gain the most from improved trauma care.

@@

Participants
------------
A total of `r exclusion[1,2]` patients were recorded in the trauma registry during the study period. Of these, `r exclusion[2,2]` were reviewed and therefore became eligible for study inclusion. `r exclusion[7,3]` patients were excluded, due to either being of paediatric age (`r exclusion[3,3]`), being dead on arrival at hospital (`r exclusion[4,3]`), or due to incomplete data (`r exclusion[5,3]`). Figure \@ref(fig:exclusion) shows the inclusion process for eligible participants.

```{r exclusion, fig.cap = "Flow diagram for application of exclusion criteria", fig.align = "center", echo=FALSE, out.width = "600px"}
knitr::include_graphics("images/exclusion.png") 
## exclusion_flowchart.R; save_png(exclusion_flowchart(exclusion), "exclusion") to update
```

Detailed totals of missing data for each required parameter in cases that were excluded for incomplete data are given in table \@ref(tab:incomplete-table). Some patients had missing data in several parameters and therefore the total number of missing parameters is greater than the total number of patients excluded due to incomplete data.

```{r incomplete-table, echo=FALSE, fig.align = "center"}
knitr::kable(table(na.data, type = "incomplete"), caption = "Parameters that were missing in cases excluded for incomplete data") %>% kable_styling(position = "center", full_width = FALSE)
```

Descriptive data
----------------
A total of `r exclusion[6,2]` participants were identified as eligible for inclusion. Their demographical and clinical characteristics are described in table \@ref(tab:table-one).

@@

```{r table-one} 
  table.one
## this thing breaks the table counting but ill just fix it manually in word
```

Main results
------------
The fitted curves for the logistic regression models of each score are shown in figure \@ref(fig:fitted-glm). The odds ratios for TRISS, NORMIT and PS were `r t[['or']]` (95% CI: `r t[['or.ci.lo']]`-`r t[['or.ci.hi']]`, *p*: `r t[['or.p']]`), `r n[['or']]` (95% CI: `r n[['or.ci.lo']]`-`r n[['or.ci.hi']]`, *p*: `r n[['or.p']]`), and `r p[['or']]` (95% CI: `r p[['or.ci.lo']]`-`r p[['or.ci.hi']]`, *p*: `r p[['or.p']]`) respectively.

```{r fitted-glm, fig.cap = "Fitted curves for the logistic regression models of each score"}
knitr::include_graphics("images/glm.png") ## update: glm_x3(df)
```

The receiver operating curves for all scores are plotted in figure \@ref(fig:roc). The areas under the curve for TRISS, NORMIT and PS were `r t[['auc']]` (95% CI: `r t[['auc.ci.lo']]`-`r t[['auc.ci.hi']]`), `r n[['auc']]` (95% CI: `r n[['auc.ci.lo']]`-`r n[['auc.ci.hi']]`), and `r p[['auc']]` (95% CI: `r p[['auc.ci.lo']]`-`r p[['auc.ci.hi']]`) respectively.

```{r roc, fig.cap = "Receiver operating curves for each score"}
knitr::include_graphics("images/roc.png") ## update: roc_x3(t, n, p)
```

The integrated calibration index for TRISS, NORMIT and PS was `r t[['ici']]` (95% CI: `r t[['ici.ci']][1]`-`r t[['ici.ci']][2]`), `r n[['ici']]` (95% CI: `r n[['ici.ci']][1]`-`r n[['ici.ci']][2]` and `r p[['ici']]` (95% CI: `r p[['ici.ci']][1]`-`r p[['ici.ci']][2]` respectively. The accuracy at all cutoffs is plotted for all scores in figure \@ref(fig:acc). The accuracy at a cutoff of 0.5 for TRISS, NORMIT and PS was `r t[['acc']]` (95% CI: `r t[['acc.ci']][1]`-`r t[['acc.ci']][2]`), `r n[['acc']]` (95% CI: `r n[['acc.ci']][1]`-`r n[['acc.ci']][2]`) and `r p[['acc']]` (95% CI: `r p[['acc.ci']][1]`-`r p[['acc.ci']][2]`) respectively.

```{r acc, fig.cap = "Accuracy at all cutoffs for each score"}
knitr::include_graphics("images/acc.png") ## update: acc_x3(t, n, p)
```

The differences between the scores for each result are shown in table \@ref(tab:delta). TRISS was superior to the other models with regards to the AUC. NORMIT was superior to the other models with regards to accuracy and to the ICI.

```{r delta, centering = TRUE}
knitr::kable(table(d, type = "deltas"), caption = "Comparison of model results") %>% kable_styling(position = "center", full_width = FALSE)
```

@@

The results of this study agree with previous studies assessing the predictive power of trauma severity models with regards to preventable deaths. Although this study was not specifically about preventable deaths, the factor that determines whether a death was preventable is whether an improvement in care could have prevented the death, i.e. whether there was a significant opportunity for improvement, which makes the previous studies on preventable deaths broadly relevant to this study. Ghorbani et al found that both TRISS and NORMIT had poor performance in predicting preventable deaths [@prevdeathsghorbani]. Heim et al found that PS and TRISS miss a significant proportion of preventable deaths [@brohi16]. Our results also show that none of the three models investigated (TRISS, NORMIT and PS) were able to reliably predict OFI. The difference between our study and prior studies is that we looked at the trauma population across the board and not only deaths.

Even though our results were expected, it was still important to investigate this question scientifically, because trauma severity models are in use today as audit filters at many institutions. Our findings show that the use of trauma severity models as audit filters is unlikely to lead to improvements in case selection, and in fact could be harmful. Audit filters that do not improve case selection will either lead to cases being over-selected, causing resource wastage, or to cases being under-selected, leading to OFI being missed. Trauma severity models are useful when used on the population level, to allow institutions to compare themselves and to provide a framework for discussing trauma severity. They are much less useful when applied to individuals, and were not intended for that use.

@@

### Health equity
<!-- Is this section requested by KI?-->
The demographical characteristics of the participants in this study reflect the well-known demographics of trauma patients worlwide. `r dd['male']`% of participants were male and the median age was `r dd['age']`, which matches with statistics both in Sweden and globally [@globalburden]. The predominance of male patients invites one to question if female patients are less likely to receive quality care. In healthcare as in all areas, repetition leads to mastery of a skill, so an onlooker might be tempted to assume that the high volume of male patients in trauma will lead healthcare providers to superiority iin treating males, whilst treatment of female patients will be neglected. This demographical difference has precipitated research that seeks to investigate whether this presumption holds true, and several large multi-centre studies have been published that evaluated whether there is a correlation between sex and outcomes. None of these studies found a correlation beween sex and increased mortality or other serious outcomes [@croce2002; @magnotti2008; @pape2019; @trentzsch2014], and in fact one study in the Netherlands found that female sex was correlated with decreased mortality [@pape2019].

### 2022-12-20 pub.Rmd
revised paper according to feedbck, wrote abstract

---
title: "Manuscript title here"
output: html_document
author: "Your name here"
bibliography: bibliography.bib
csl: vancouver.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- This is an R Markdown document. Markdown is a simple formatting
syntax for authoring HTML, PDF, and MS Word documents. For more
details on using R Markdown see http://rmarkdown.rstudio.com -->

<!-- This is a comment and will not be present in the compiled
document. Try "knitting" this document by clicking the knit button up
to the left, or by running `rmarkdown::render("manuscript.Rmd")` in
the console and you should only see headings -->

<!-- Structure and most comments are from the Strengthening the
reporting of observational studies in epidemiology (STROBE) statement,
see https://www.strobe-statement.org/. For more explanations see the
paper Strengthening the Reporting of Observational Studies in
Epidemiology (STROBE):explanation and elaboration.  -->

<!-- Indicate the study’s design with a commonly used term in the
title, for example: "Associations between resuscitation interventions
and opportunities for improvement in adult trauma patients: A cohort
study" -->

>>>>

---
title: "Association between trauma severity models and opportunities for improvement:
  A retrospective cohort study"
output: html_document
author: "Victoria Witthoff"
bibliography: bibliography.bib
csl: vancouver.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r main}
  ## source("main.R") 
```
